---
title: "Homework 2 zz2901"
author: "Troy Zhou"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: github_document
---

```{r load_package, echo = FALSE, message=FALSE}
library(tidyverse)
options(tibble.print_min = 5)
```

## Problem 1

Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7.

```{r}
trans_ent = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```

The next code chunk is similar, but filters according to ADA compliance as an initial step. This produces a dataframe in which the number of rows is the number of ADA compliant stations. 

```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion (recall that R will coerce logical to numeric in cases like this).

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

Lastly, we write a code chunk to identify stations that serve the A train, and to assess how many of these are ADA compliant. As a first step, we tidy the data as alluded to previously; that is, we convert `route` from wide to long format. After this step, we can use tools from previous parts of the question (filtering to focus on the A train, and on ADA compliance; selecting and using `distinct` to obtain dataframes with the required stations in rows).

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct

trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

## Problem 2

read and clean Mr. Trash Wheel Sheet

```{r}
mr_trash_wheel = 
  readxl::read_excel("data/Trash Wheel Collection Data.xlsx", sheet="Mr. Trash Wheel",
                     range = "A2:N550") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster) & !is.na(month)) %>% 
  mutate(sheet="Mr. Trash Wheel",year=as.numeric(year),sports_balls=as.integer(round(sports_balls))) %>% 
  select(sheet,everything())
```

read and clean Professor Trash Wheel Sheet

```{r}
prof_trash_wheel = 
  readxl::read_excel("data/Trash Wheel Collection Data.xlsx", sheet="Professor Trash Wheel", 
                     range = "A2:M97") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster) & !is.na(month))%>% 
  mutate(sheet="Professor Trash Wheel") %>% 
  select(sheet,everything())
```

combine the two data sets

```{r}
mr_prof_trash_wheel =
  full_join(mr_trash_wheel,prof_trash_wheel)
mr_prof_trash_wheel
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables.

* The data sets are both imported using omitting the total rows and with new variable called `sheet` to record their sheet of origin. The variable names are cleaned using `janior` package's `clean_names()` function. 

* The two data sets have same variables, `ncol(mr_prof_trash_wheel)` except that `prof_trash_wheel` missed the `sports_balls`. `dumpster` is the ID for the dumpster. `month`, `year`, `date` are date variables. `weight_tons`:`homes_powered` are variables recording the corresponding data for each dumpster. 

* `mr_trash_wheel` contains ``r nrow(mr_trash_wheel)`` observations. And `prof_trash_wheel` contains ``r nrow(prof_trash_wheel)`` observations. It was created by using `full_join()`

* The combined data set `mr_prof_trash_wheel` contains ``r nrow(mr_prof_trash_wheel)`` observations.

For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of sports balls collected by Mr. Trash Wheel in 2020?

* The total weight of trash collected by Professor Trash Wheel was ``r sum(mr_prof_trash_wheel$weight_tons[which(mr_prof_trash_wheel$sheet=="Professor Trash Wheel")])`` tons.

* The total number of sports balls collected by Mr. Trash Wheel in 2020 was ``r sum(mr_prof_trash_wheel$sports_balls[which(mr_prof_trash_wheel$sheet=="Mr. Trash Wheel" & mr_prof_trash_wheel$year==2020)])``.


** Problem 3

First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols_month = read_csv("data/fivethirtyeight_datasets/pols-month.csv") %>% 
  separate(mon,sep="-",c("year","month","day")) %>%
  arrange(year,month) %>% 
  mutate(year=as.numeric(year),month=month.abb[as.numeric(month)], president=recode(prez_dem,`0`="gop",`1`="dem"))  %>% 
  select(year,month,everything(),-prez_dem,-prez_gop,-day)  
  
pols_month
```


Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp = read_csv("data/fivethirtyeight_datasets/snp.csv")%>% 
  mutate(date=as.Date(date,"%m/%d/%y")) %>% 
  separate(date,sep="-",c("year","month","day")) %>%
  arrange(year,month) %>% 
  mutate(year=as.numeric(year),month=month.abb[as.numeric(month)])  %>% 
  select(year,month,everything(), -day)  
snp
```


Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
unemployment = read_csv("data/fivethirtyeight_datasets/unemployment.csv") %>% 
  rename(year=Year) %>% 
  mutate(year=as.numeric(year)) %>% 
  arrange(year) %>% 
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment_p"
  ) 
  
unemployment

```


Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
pols_snp_merge = 
  full_join(pols_month,snp,by=c("year","month"))
pols_snp_merge

pols_snp_unemployment = 
  full_join(pols_snp_merge,unemployment,by=c("year","month"))
pols_snp_unemployment
```


Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

* Data set `pols_month` after tidying contains ``r nrow(pols_month)`` observations and ``r ncol(pols_month)`` variables, including results from ``r pols_month$year[1]`` to ``r pols_month$year[nrow(pols_month)]``. 

  * Variable `president` was recoded from `prez_dem` and `prez_gop`. It indicates whether the president of the corresponding year is democratic or republican. Rest of variables start with `gov_`, `sen_`, `rep_`` indicates the number of governors, senators, or representatives for each party in that year.
  
* Data set `snp` contains `date` and `close` variables. After tidying and seperating `date` into `year` and `month` variables, it has ``r nrow(snp)`` observations and ``r ncol(snp)`` variables.

  * Variable `close` indicates the closing values of the S&P stock index on the associated date.
  
* Data set `unemployment` originally contains `Year` and percentage of unemployment percentage for each month in wide format. After transforming it into long format, it has ``r nrow(unemployment)`` observations and ``r ncol(unemployment)`` variables.

  * Variable `month` stores the original column names, which represents months. 
  * Variable `unemployment_p` stores the row values for each month's unemployment precentages.

